---
title: 'Introductory Statistics with Randomization and Simulation: Chapter 4'
output:
   ioslides_presentation:
     font-family: Lato Semibold
     font-import: http://fonts.googleapis.com/css?family=Lato
     widescreen: yes
     css: ../style.css
     fig_caption: yes
---
<style>
citation {
  font-size: 4px;
}
</style>

<!--  Version 1.0-0

      This version of the slides is taken directly from Mine Çetinkaya-Rundel's lecture slides
      posted on OpenIntro.org in .pptx and .gdslides format. Simply moved to Rmd. 

      A large part of the HTML/CSS formatting is janky, and could be cleaned up. Feel free to issue a 
      pull request if you love HTML and CSS and want to fix this up.

      - wburr, Oct 16, 2017
-->

<!-- This is Chapter 4.1 in the text, slides by Mine Cetinkaya-Rundel -->


<!-- This is Chapter 4.5 in the text, slides by Wesley Burr, based on text -->

# The Bootstrap

## Quantifying Uncertainty

The theory required to quantify uncertainty of sample standard deviations
is quite complex.

* Ideally, we would sample data from the population, again and again
* Repeat until we have enough standard deviation estimates that we can determine precision of original estimate
* This is **very expensive**

## The Bootstrap

Since it is expensive to resample from the population, the bootstrap (invented by Bradley Efron in 1979) resamples **from the original sample**.

Given an original sample of size $N$:
* Randomly sample one observation from the set of size $N$
* Randomly sample a second observation from the set of size $N$
* $\cdots$
* Randomly sample an $N^\text{th}$ observation from the set of size $N$

This, as we've seen before (when doing permutation simulations) is known as **sampling with replacement**. 

## The Bootstrap (ctd.)

Once we have a new, sampled with replacement, sample, we compute our test statistic
again on the new sample. 

Iterate this process
* Sample with replacement $N$ times, compute the test statistic
* Do the entire process again
* And again ...

This gives the **bootstrap distribution** of the test statistic. 

## Example: Amazon Book Prices

```{r, echo = FALSE}
load("textbooks.rda")
```

```{r, eval = FALSE}
data(textbooks, package = "openintro")
sd(textbooks$diff)
```

```{r, echo = FALSE}
round(sd(textbooks$diff), digits = 3)
```
This is the data set from Chapter 4.2: some prices on textbooks from Amazon.com
and a university bookstore. We have 73 observations of textbook prices, with standard deviation `r round(sd(textbooks$diff), digits = 3)`.

## Example (ctd.)

We can sample from this data set of 73 differences, repeatedly.

```{r, eval = FALSE}
sd_vec <- vector(length = 10000)
for(j in 1:10000) {
  sd_vec[j] <- sd(sample(textbooks$diff, size = 73, replace = TRUE))
}
hist(sd_vec)
```

<center>
```{r, echo = FALSE, fig.height = 3}
set.seed(69)
sd_vec <- vector(length = 10000)
for(j in 1:10000) {
  sd_vec[j] <- sd(sample(textbooks$diff, size = 73, replace = TRUE))
}
hist(sd_vec, breaks = seq(6, 22, 0.5))
```
</center>

## Example (ctd.)

The bootstrap distribution is symmetric, bell-shaped, and centered at `r round(mean(sd_vec), digits = 3)`, near the standard deviation of the textbook data, `r round(sd(textbooks$diff), digits = 3)`. 

* The standard deviation of the bootstrap deviation is `r round(sd(sd_vec), digits = 3)`.
* We can then use this to quantify the uncertainty of the point estimate
* This is the **bootstrap estimate** of the **standard error** (SE)

## Conditions for using Bootstrap for Inference

If 

* the observations in the original sample are independent
* the original sample size $N \geq 30$
* the bootstrap distribution is approximately normal

then the bootstrap distribution for the standard deviation will be a good approximation of the sampling distribution for the standard deviation.

## Third Condition

How do we check that the bootstrap distribution is approximately normal? **Use a Q-Q plot**!

```{r,eval = FALSE}
qqnorm(sd_vec)
```
<center>
```{r, fig.height = 3.5, echo = FALSE}
qqnorm(sd_vec)
```
</center>

## Confidence Interval

Once we have the conditions met, we can do standard techniques that require normality: compute confidence intervals, perform inference, etc.

We use the same formula as before:

$$
\text{point estimate} \pm  t_{df}^* \cdot \text{SE}
$$

Since we have $N = 73$, this gives $df = N-1 = 72$. Then,
```{r}
qt(0.975, df = 72)
```

## Confidence Interval

We then compute a 95% confidence interval for the standard deviation of the differences in textbook prices:

$$
\begin{split}
\text{point estimate} &\pm  t_{df}^* \cdot \text{SE} \\
14.26 & \pm 1.9935 \cdot 1.587 \\
(11.096, &17.424)
\end{split}
$$

So we say we are 95% confident that the standard deviation of the textbook price differences is between \$11.10 and \$17.42.

## Hypothesis Tests

We can also use this method for the standard error (SE) to compute a hypothesis test. 

## Applicability of Bootstrap

In general, we can use the bootstrap method for other parameters, using the same conditions as we had earlier. However, be careful: **always check** the third condition, for normality of the bootstrap distribution.
